draft

### 2025-05-21 draft.tex
Updates from Overleaf

\subsection{Engineered features in the era of deep learning}
\label{sec:org9dc3dfa}
An alternative to extracting a defined set of engineered features is to instead use a trained deep learning model to extract features. Results are mixed, with suitably-trained deep learning networks usually outperforming engineered features, but not always \citep{tangMorphologicalProfilingDrug2024}. Training a deep learning model to be a good feature extractor takes significant hands-on and compute time, and often requires appropriate training images. Trained models can in some cases be used on new datasets, eliminating the training step. Generally deep learning-based feature extraction takes less compute time than CellProfiler does; however, the features are not obviously interpretable and require specialized techniques for biologists to attempt to understand patterns of changes in image-based profiles, which is not always successful \citep{moenDeepLearningCellular2019}. Compared to engineered features, deep learning approaches have a higher barrier of entry, as most models require annotating data. Foundational models, whose goal is to lower this barrier, fail to generalize well for out-of-distribution data. Unlike learned features, whose interpretation requires complex methods like counterfactual methods or attention maps, most engineered features have a mathematical definition that is faster to calculate and easier to translate into biological concepts.

>>>>

\subsection{Engineered features in the era of deep learning}
\label{sec:org9dc3dfa}
An alternative to extracting a defined set of engineered features is to instead use a trained deep learning model to extract features. Results are mixed, with suitably-trained deep learning networks usually outperforming engineered features \cite{lafargeCapturingSingleCellPhenotypic2019,moshkovLearningRepresentationsImagebased2022,chowPredictingDrugPolypharmacology2022,wolfSCANPYLargescaleSinglecell2018}, but not always \cite{tangMorphologicalProfilingDrug2024,kimSelfsupervisionAdvancesMorphological2023}. Training a deep learning model to be a good feature extractor takes significant hands-on and compute time, and often requires appropriate training images. Trained models can in some cases be used on new datasets, eliminating the training step. Compared to engineered features, deep learning approaches have a higher barrier of entry, as most models require annotating data. Foundational models, whose goal is to lower this barrier, fail to generalize well for out-of-distribution data \cite{aza}. Generally deep learning-based feature extraction takes less compute time than CellProfiler does; however, the features are not obviously interpretable and require specialized techniques for biologists to attempt to understand patterns of changes in image-based profiles, which is not always successful \citep{moenDeepLearningCellular2019}.  Unlike learned features, whose interpretation requires complex methods like counterfactual methods or attention maps, most engineered features have a mathematical definition that is faster to calculate and easier to translate into biological concepts.
